---
title: "(Cross-) Recurrence Quantification Analysis"
author: "Fred Hasselman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{(Cross-) Recurrence Quantification Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown} 
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	fig.align = "center",
	fig.width = 7,
	message = FALSE,
	warning = FALSE,
	collapse = FALSE,
	comment = ">",
	dpi = 72,
	width = 500
)
library(invctr)
```


# CRQA

There are 2 main ways to run CRQA in `casnet`:

* Using functions `rp`, `rp_measures` and `rp_plot`
* Using function `rp_cl` which will run Norbert Marwan's [commandline Recurrence Plots](http://tocsy.pik-potsdam.de/commandline-rp.php)

The following example will use the native `casnet` functions, see the paragraph **An R interface to Marwan's commandline recurrence plots** to learn about using `rp_cl()`.



## **Categorical Auto RQA**

We'll use data from @oomens2015 in which 242 students were asked to generate random sequences of 100 numbers between 1 and 9 (for details see [the article](https://www.frontiersin.org/articles/10.3389/fnhum.2015.00319/full)).

```{r RNG1, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(casnet)
library(ggplot2)

# Load the random number sequence data from Oomens et al. (2015)
data(RNG)

# Select a subject
IDs <- RNG$ID%in%c(163,291)

# Look at the sequence
ggplot(RNG[IDs,],aes(x=time,y=number,group=ID)) +
  geom_line(aes(colour=ID))+
  facet_grid(~ID) +
  scale_y_continuous(breaks = 1:9) +
  ggtitle("Which of these number sequences is more 'random'?") +
  theme_bw()
```

In order to answer the question in the Figure title, we'll run a Recurrence Quantification Analysis.

The data are unordered categorical, that is, the differences between the integers are meaningless in the context of generating random number sequences. This means the RQA parameters can be set to quantify recurrences of the same value: 

* Embedding lag = 1
* Embedding dimension = 1
* Radius = 0 (any number $\leq$ 1 will do)

In the code block below the functions `rp()`, `rp_measures()` and `rp_plot()` are used to perform RQA on 2 participants in the dataset.

```{r RNG2, message=FALSE, warning=FALSE, collapse=FALSE, include=TRUE, echo=TRUE, paged.print=FALSE}
# Run the RQA analysis
y_1  <- RNG$number[RNG$ID==163]
y_2 <-  RNG$number[RNG$ID==291]

## Plot the recurrence matrix
# Get the recurrence matrix
rp_1 <- rp(y1=y_1, emDim = 1, emLag = 1, emRad = 1)
rp_2 <- rp(y1=y_2, emDim = 1, emLag = 1, emRad = 1)

# Get the plots
g_1 <- rp_plot(rp_1, plotDimensions = TRUE, returnOnlyObject = TRUE, title = "ID 163")
g_2 <- rp_plot(rp_2, plotDimensions = TRUE, returnOnlyObject = TRUE, title = "ID 291")

# Get the RQA measures, using silent = FALSE will produce output in the console.
crqa_1 <- rp_measures(rp_1, silent = FALSE)
crqa_2 <- rp_measures(rp_2, silent = FALSE)

# Using rp_cl() would look very similar:
# rp_1 <- rp_cl(y1 = y_1, emDim  = 1, emLag = 1, emRad= 1)
# rp_2 <- rp_cl(y1 = y_2, emDim  = 1, emLag = 1, emRad= 1)
```


The output from `rp_measures()` is structured into _Global_ and _Line-based_ measures. 
In addition to information about the matrix size and number of points global measures are provided such as the _Recurrence Rate_ (RR), the number of points that do **not** form a line (Singular Points) the _Divergence_ (1 / the longest line structure) the average _Repetitiveness_ (proportion of horizontal and vertical lines on the number of diagonal lines). The idea is to quantify how much of the deterministic structure (line structures) is due to repeating the same values (i.e., the horizontal and vertical lines, Laminarity). Finally the _Anisotropy_ quantifies the symmetry of line structures in the plot by taking the ratio of he number of vertical lines over the numbers of horizontal lines in the plot. This should be 1 for Auto-RQA.

The _Line-based_ output is a table listing the statistics for diagonal, vertical and horizontal lines (mean length, max length, rate, entropy of the distribution of line lengths and the same entropy but relative to the number of possible recurrent points and the coefficient of variation of line lengths).

The actual output object is a dataframe (which has more output fields, see the manual pages), the table output to the console is added as attribute `measuresTable`,

Below the data and plots are rearranged for ease of comparison.

```{r RNG3, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(cowplot)
# The recurrence plots
cowplot::plot_grid(g_1, g_2)

# The RQA measures
cbind.data.frame(subj163=t(crqa_1), subj291=t(crqa_1))

# The tables are stored in an attribute
attr(crqa_1,"measuresTable")
attr(crqa_2,"measuresTable")
```

The sequence generated by participant 163 has a higher **DET**erminism (`DET = .40`) than the sequence by particpant 291 (`DET = .19`). The ratio of points on a diagonal line to the total number of recurrent point also quantifies this difference (`DET_RR`). Also interesting to note, both participants have a **LAM**inarity score of `0`. This implies they avoided to produce patterns in which the exact same numbers were repeated in succession. This is a tell-tale sign of the *non-random* origins of these sequences.


## **Hypothesis testing using constrained data realisations** 

A simple strategy to get some more certainty about the differences between the two sequences is to randomise the observed series, thus removing any temporal correlations that might give rise to recurring patterns in the sequences and re-run the RQA. If the repeated patterns generated by participant 163 are non-random one would expect the **DET**erminism to drop. If they do not drop this could indicate some random autoregressive process is causing apparent deterministic temporal patterns.

```{r RNG4, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Reproduce the same randomisation
set.seed(123456789)

# Randomise the number sequences
y_1rnd <- y_1[sample(1:NROW(y_1),size = NROW(y_1))]
y_2rnd <- y_2[sample(1:NROW(y_2),size = NROW(y_2))]

# Create the recurrence matrix
rp_1rnd <- rp(y1=y_1rnd, emDim = 1, emLag = 1,emRad = 1)
rp_2rnd <- rp(y1=y_2rnd, emDim = 1, emLag = 1,emRad = 1)

# Get the RPs
g_1rnd <- rp_plot(rp_1rnd, plotDimensions = TRUE, returnOnlyObject = TRUE, title = "ID 163 shuffled")
g_2rnd <- rp_plot(rp_2rnd, plotDimensions = TRUE, returnOnlyObject = TRUE, title = "ID 291 shuffled")

# Get CRQA measures
crqa_1rnd <- rp_measures(rp_1rnd, silent = FALSE)
crqa_2rnd <- rp_measures(rp_2rnd, silent = FALSE)

# Display recurrence plots
cowplot::plot_grid(g_1rnd, g_2rnd, align = "h")

# Display the RQA measures for ID 163
cbind.data.frame(subj163=t(crqa_1), subj163rnd=t(crqa_1rnd))

# Display the RQA measures for ID 291
cbind.data.frame(subj291=t(crqa_2),  subj291rnd=t(crqa_2rnd))
```

Note that the number of recurrent points (`RR`) does not change whe we shuffle the data. What changes is the number of recurrent points that form line structures in the recurrence plot. Randomising the number sequences causes vertical line structures to appear in the recurrence plot (`LAM`, `V_max`, `V_entr`, `TT`), this is what we would expect if the data generating process were indeed a random process. Having no such structures means there were hardly any sequences consisting of repetitions of the same number. Participants may have adopted a strategy to avoid such sequences because they erroneously believed this to be a feature of non-random sequences.


### A permutation test with surrogate time series
In order to get an idea about the meaningfulness of these differences, we can construct a surrogate data test for each participant. If we want a one-sided test with $\alpha=.05$, the formula for the number of constrained realisations $M$ we minimally need is: $$M = \frac{1}{\alpha}-1 = 19$$. Add the observed value and we have a sample size of $N = 20$. For a two sided test we would use $$M = \frac{2}{\alpha}-1 = 39$$.    

Of course, if there are no computational constraints on generating surrogate time series, we can go much higher, If we want $N = 100$, the test will be an evaluation of $H_{0}$ at $\alpha = .01$.

1. Create `99` realisations that reflect a test of the hypothesis $H_{0}: X_i \sim \mathcal{U(1,9)}$ at $\alpha = .01$.
2. Calculate the measure of interest, e.g. `DET`
3. If the observed `DET` value is at the extremes of the distribution of values representing $H_{0}$, the observed value was probably not generated by drawing from a discrete uniform distribution with finite elements `1` through `9`.

```{r RNG5, echo=TRUE, message=FALSE, warning=FALSE,  include=FALSE, paged.print=FALSE}
library(plyr)
library(dplyr)

set.seed(123456789)

y_1rnd_sur <- ldply(1:99, function(s) y_1[sample(1:NROW(y_1),size = NROW(y_1))])
y_2rnd_sur <- ldply(1:99, function(s) y_2[sample(1:NROW(y_2),size = NROW(y_2))])

# If returnMeasures = TRUE the outut from rp_measures() will be saved in an attribute "measures".
crqa_1rnd_sur <- ldply(seq_along(y_1rnd_sur$V1), function(r){
  tmp <- rp(y1 = as.numeric(y_1rnd_sur[r,]), emDim  = 1, emLag = 1, emRad= 1, returnMeasures = TRUE)
  return(attr(tmp,"measures"))
  })
crqa_1rnd_sur[NROW(crqa_1rnd_sur)+1,] <- crqa_1
crqa_2rnd_sur <- ldply(seq_along(y_2rnd_sur$V1), function(r){
  tmp <- rp(y1 = as.numeric(y_2rnd_sur[r,]), emDim  = 1, emLag = 1, emRad= 1, returnMeasures = TRUE)
  return(attr(tmp,"measures"))
  })
crqa_2rnd_sur[NROW(crqa_2rnd_sur)+1,] <- crqa_2
```

Use function `plorSUR_hist()` to get a p-value and plot the distributions. The red dots indicate the observed values.

```{r RNG5a, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.height=8}
# Get point estimates for p-values based on rank of observation (discrete distribution)
#   99 = (1 / alpha) - 1
# 99+1 = (1 / alpha)
alpha = 1/100

p_1 <- plotSUR_hist(surrogateValues = crqa_1rnd_sur$DET, observedValue = crqa_1$DET, measureName = "DET", doPlot = FALSE)
p_2 <- plotSUR_hist(surrogateValues = crqa_2rnd_sur$DET, observedValue = crqa_2$DET, measureName = "DET", doPlot = FALSE)

cowplot::plot_grid(p_1$surrogates_plot, p_2$surrogates_plot, labels = c("ID 163","ID 291"), ncol = 1)
```


To get the full picture, let's look at those missing repetitions of the same numbers.

```{r RNG6, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.height=8}
# Get point estimates for p-values based on rank of observation (discrete distribution)
#   99 = (1 / alpha) - 1
# 99+1 = (1 / alpha)
alpha = 1/100

p_1 <- plotSUR_hist(surrogateValues = crqa_1rnd_sur$LAM_vl, observedValue = crqa_1$LAM_vl, measureName = "LAM", doPlot = FALSE)
p_2 <- plotSUR_hist(surrogateValues = crqa_2rnd_sur$LAM_vl, observedValue = crqa_2$LAM_vl, measureName = "LAM", doPlot = FALSE)

cowplot::plot_grid(p_1$surrogates_plot, p_2$surrogates_plot, labels = c("ID 163","ID 291"), ncol = 1)
```

If we were naive to the origin of these number sequences, the results for **LAM**inarity should make us doubt that they represent indendent draws from a discrete uniform distribution of the type $X \sim \mathcal{U}(1,9)$. If we had to decide which sequence was more, or, less random, then based on the **DET**erminism result, we would conclude that participant 163 produced a sequence that is less random than participant 291, the observed value of the former is at the right extreme of a distribution of `DET` values calculated from 99 realisations of the data constrained by $H_0$.


## Categorical Cross-RQA



## Continuous Auto-RQA



## Continuous Cross-RQA






### **An R interface to Marwan's commandline recurrence plots**

> **IMPORTANT**: Currently `rp_cl` can only run on an operating system that allows execution of 32-bit applications!

The `rp_cl()` function is a wrapper for the [commandline Recurrence Plots](http://tocsy.pik-potsdam.de/commandline-rp.php) executable provided by Norbert Marwan. 

The `rp` executable is installed on your machine when the function `rp_cl()` is called for the first time:

* It is renamed to `rp` from a platform specific file downloaded from the [commandline Recurrence Plots](http://tocsy.pik-potsdam.de/commandline-rp.php) site.
* The file is copied to the directory: `r normalizePath("[path to casnet]/exec/",mustWork = FALSE)`
   - Make sure that you have rights to execute programs in this directory!
* The latter location is stored as an option and can be read by calling `getOption("casnet.path_to_rp")`

If you cannot change the permissions on the folder where `rp` was downloaded, consider downloading the appropriate executable from the [commandline Recurrence Plots](http://tocsy.pik-potsdam.de/commandline-rp.php) site to a directory in which you have such permissions. Then change the `path_to_rp` option using `options(casnet.path_to_rp="YOUR_PATH_TO_RP")`. See the manual entry for `rp_cl()` for more details.

---

The platform specific `rp` command line executable files were created by Norbert Marwan and obtained under a Creative Commons License from the website of the Potsdam Institute for Climate Impact Research at: http://tocsy.pik-potsdam.de/


The full copyright statement on the website is as follows:    

> © 2004-2017 SOME RIGHTS RESERVED    
> University of Potsdam, Interdisciplinary Center for Dynamics of Complex Systems, Germany    
> Potsdam Institute for Climate Impact Research, Transdisciplinary Concepts and Methods, Germany    
> This work is licensed under a [Creative Commons Attribution-NonCommercial-NoDerivs 2.0 Germany License](https://creativecommons.org/licenses/by-nc-nd/2.0/de/).    

More information about recurrence quantification analysis can be found on the [Recurrence Plot website](http://www.recurrence-plot.tk).


## **References**
