---
title: "Recurrence Networks"
author: "Fred Hasselman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recurrence Networks}
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
	warning = FALSE,
  comment = ">",
  paged.print = TRUE,
  fig.align = "center",
  fig.width = 7,
  fig.height=7,
  width = 500,
  dpi=72
)
library(casnet)
library(igraph)
library(kableExtra)
```

# *Recurrence Networks*

A recurrence network is a representation of a recurrence matrix as a graph (a network). The nodes in the network represent time points, and if a value at any point in time will recur at some later point in time, an edge is drawn between the time points.


## Recurrence Networks 

Recurrence networks are graphs created from a recurrence matrix. This means the nodes of the graph represent time points and the connections between nodes represent a recurrence relation betwen the values observed at those time points. That is, often the matrix represents recurrences in a reconstructed state space, the values are coordinates and therefore we would say the edges of a recurrence network represent a temporal relation between recurring states. The ultimate reference for learning everything about recurrence networks is:

> [Zou, Y., Donner, R. V., Marwan, N., Donges, J. F., & Kurths, J. (2018). Complex network approaches to nonlinear time series analysis. Physics Reports](https://www.sciencedirect.com/science/article/abs/pii/S037015731830276X?via%3Dihub)


Package `casnet` has some functions to create recurrence networks, they are similar to the functions used for CRQA:
* `rn()` is very similar to `rp()`, it will create a matrix based on embedding parameters. One difference is the option to create a weighted matrix. This is a matrix in which non-recurring values are set to 0, but the recurring values are not replaced by a 1, the distance value is retained and acts as an edge-weight
* `rn_plot()` will produce the same as `rp_plot()`

We can turn the recurrence matrix into an adjacency matrix, an `igraph` object. This means we can use all the `igraph` functions to calculate network measures.

```{r}
library(igraph)
#library(qgraph)
#library(survival)
# Reload the data we used earlier
series <- rio::import("https://github.com/complexity-methods/CSA-assignments/raw/master/assignment_data/BasicTSA_arma/series.xlsx")

# Lets use a shorter dataset to speed things up
series <- series[1:500,]
```

We'll analyse the three time series as a recurrence network:
* Compare the results by comparing the different measures
   - Remember: TS_1 was white noise, TS_2 was a sine with added noise, TS_3 was the logistic map in the chaotic regime.
* Note that some of the RQA measures can be exactly calculated from the measures of the network representation.
   - Try to understand why the Recurrence is represented as the *degree centrality* of the network (`igraph::centr_degree()`)


### TS 1 {-}

```{r message=FALSE, warning=FALSE}
#----------------------
# Adjacency matrix TS_1
#----------------------
plot(ts(series$TS_1))

arcs <- 6

# Because these are generated signals, look for a drop in FNN below 1%.
p1 <- est_parameters(y = series$TS_1, nnThres = 1)

# By passing emRad = NA, a radius will be calculated
RN1 <- casnet::rn(y1 = series$TS_1, emDim = p1$optimDim, emLag = p1$optimLag, emRad = NA, targetValue = 0.05)
casnet::rn_plot(RN1)

# Get RQA measures
rqa1 <- casnet::rp_measures(RN1, silent = FALSE)

# Create RN graph
g1  <- igraph::graph_from_adjacency_matrix(RN1, mode="undirected", diag = FALSE)
igraph::V(g1)$size <- igraph::degree(g1)
g1r <- casnet::make_spiral_graph(g1,arcs = arcs, epochColours = getColours(arcs), markTimeBy = TRUE)

# Get RN measures
rn1 <- rn_measures(g1, silent = FALSE)

# Should be the same
rqa1$RR==rn1$graph_measures$EdgeDensity
```

### TS 2 {-}

```{r message=FALSE, warning=FALSE}
#----------------------
# Adjacency matrix TS_2
#----------------------
plot(ts(series$TS_2))

# Because these are generated signals, look for a drop in FNN below 1%.
p2 <- est_parameters(y = series$TS_2, nnThres = 1)

RN2 <- rn(y1 = series$TS_2, emDim = p2$optimDim, emLag = p2$optimLag, emRad = NA,  targetValue = 0.05)
rn_plot(RN2)

# Get RQA measures
rqa2 <- rp_measures(RN2, silent = FALSE)

# Create RN graph
g2 <- igraph::graph_from_adjacency_matrix(RN2, mode="undirected", diag = FALSE)
V(g2)$size <- degree(g2)
g2r <- make_spiral_graph(g2,arcs = arcs ,epochColours = getColours(arcs), markTimeBy = TRUE)

# Get RN measures
rn2 <- rn_measures(g2, silent = FALSE)

# Should be the same
rqa2$RR==rn2$graph_measures$EdgeDensity

```

### TS 3 {-}

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
#----------------------
# Adjacency matrix TS_3
#----------------------
plot(ts(series$TS_3))

# Because these are generated signals, look for a drop in FNN below 1%.
p3 <- est_parameters(y = series$TS_3, nnThres = 1)

RN3 <- rn(y1 = series$TS_3, emDim = p3$optimDim, emLag = p3$optimLag, emRad = NA, targetValue = 0.05)
rn_plot(RN3)

# Get RQA measures
rqa3 <- rp_measures(RN3, silent = FALSE)

# Create RN graph
g3 <- igraph::graph_from_adjacency_matrix(RN3, mode="undirected", diag = FALSE)
V(g3)$size <- degree(g3)
g3r <- make_spiral_graph(g3,arcs = arcs ,epochColours = getColours(arcs), markTimeBy = TRUE)

# Get RN measures
rn3 <- rn_measures(g3, silent = FALSE)

# Should be the same
rqa3$RR==rn3$graph_measures$EdgeDensity
```


## Multiplex Recurrence Networks 

Consider the three time series to be part of a multi-layer recurrence network.
Common properties of the multiplex network are _inter-layer mutual information_ and _edge overlap_ can be calculated using function `casnet::mrn()`. One problem, the networks have to be all of the same size (same number of nodes, a multivariate time series), but here we have reconstructed the phase space using different embedding parameters... let's choose one set of parameters for all time series. 

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
emDim <- mean(c(p1$optimDim,p2$optimDim,p3$optimDim))
emLag <- median(c(p1$optimLag,p2$optimLag,p3$optimLag))

RNs <- plyr::llply(1:3, function(r) rn(y1 = series[,r], emDim = emDim, emLag = emLag, emRad = NA, targetValue = 0.05))
layers <- plyr::llply(RNs, function(r) igraph::graph_from_adjacency_matrix(r, mode="undirected", diag = FALSE))
names(layers) <- c("g1","g2","g3")

mrn(layers = layers)
```

A variety of plots can be created using `casnet::mrn_plot()`

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}

# Simple
mrn_plot(layers = layers, showEdgeColourLegend =TRUE)
mrn_plot(layers = layers, MRNweightedBy = "EdgeOverlap", showEdgeColourLegend =TRUE)


# Include picture of Layers 
mrn_plot(layers = layers, RNnodes = TRUE)
mrn_plot(layers = layers, RNnodes = TRUE,MRNweightedBy = "EdgeOverlap", showEdgeColourLegend =TRUE)
```
  
## Time-varying Multiplex Recurrence Networks 

The MRN can be calculated in a sliding window, by setting the arguments `win` and `step`.

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}

# This will generate 26 windows
MRN_win <- mrn(layers = layers, win = 250, step = 10)

# The MRN are returned as a list
MRN_win$interlayerMI

# It may be informative to create an animation [not shown here because it displays in the Viewer]
# MRN_ani <- mrn_plot(layers = layers, win = 250, step = 10, createAnimation = TRUE)

# The animation is stored in an output field, but is also saved as a .gif, see the man pages for more options.
# MRN_ani$MRNanimationGG

```
  