---
title: "Recurrence Networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RecurrenceNetworks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  paged.print = TRUE,
  fig.align = "center",
  fig.width = 7,
  fig.height=7,
  width = 500
)
library(casnet)
library(igraph)
library(kableExtra)
```

# *Recurrence Networks*

A recurrence network is a representation of a recurrence matrix as a graph (a network). The nodes in the network represent time points, and if a value at any point in time will recur at some later point in time, an edge is drawn between the time points.


## Recurrence Networks {.tabset .tabset-fade .tabset-pills}

Recurrence networks are graphs created from a recurrence matrix. This means the nodes of the graph represent time points and the connections between nodes represent a recurrence relation betwen the values observed at those time points. That is, often the matrix represents recurrences in a reconstructed state space, the values are coordinates and therefore we would say the edges of a recurrence network represent a temporal relation between recurring states. The ultimate reference for learning everything about recurrence networks is:

> [Zou, Y., Donner, R. V., Marwan, N., Donges, J. F., & Kurths, J. (2018). Complex network approaches to nonlinear time series analysis. Physics Reports](https://www.sciencedirect.com/science/article/abs/pii/S037015731830276X?via%3Dihub)


Package `casnet` has some functions to create recurrence networks, they are similar to the functions used for CRQA:
* `rn()` is very similar to `rp()`, it will create a matrix based on embedding parameters. One difference is the option to create a weighted matrix. This is a matrix in which non-recurring values are set to 0, but the recurring values are not replaced by a 1, the distance value is retained and acts as an edge-weight
* `rn_plot()` will produce the same as `rp_plot()`

We can turn the recurrence matrix into an adjecency matrix, an `igraph` object. This means we can use all the `igraph` functions to calculate network measures.

```{r}
library(igraph)
library(qgraph)
# Reload the data we used earlier
series <- rio::import("https://github.com/complexity-methods/CSA-assignments/raw/master/assignment_data/BasicTSA_arma/series.xlsx")

# Lets use a shorter dataset to speed things up
series <- series[1:500,]
```

We'll analyse the three time series as a recurrence network:
* Compare the results, look at the SMall World Index and other measures
   - Remember: TS_1 was white noise, TS_2 was a sine with added noise, TS_3 was the logistic map in the chaotic regime.
* Note that some of the RQA measures can be exactly calculated from the measures of the network representation.
   - Try to understand why the Recurrence is reprented as the *degree centrality() of the network (`igraph::centr_degree()`)


### TS 1 {-}

```{r message=FALSE, warning=FALSE}
#----------------------
# Adjacency matrix TS_1
#----------------------
plot(ts(series$TS_1))

arcs=6

est_parameters(y = series$TS_1)

# By passing emRad = NA, a radius will be calculated
RN1 <- casnet::rn(y1 = series$TS_1, emDim = 7, emLag = 2, emRad = NA, targetValue = .01)
casnet::rn_plot(RN1)

# Get RQA measures
rqa1 <- rp_measures(RN1)

# knitr::kable(rqa1,digits = 2) %>%
#   kable_styling() %>%
#   scroll_box(width = "100%", height = "100px")

g1  <- igraph::graph_from_adjacency_matrix(RN1, mode="undirected", diag = FALSE)
igraph::V(g1)$size <- igraph::degree(g1)
g1r <- casnet::make_spiral_graph(g1,arcs = arcs, epochColours = getColours(arcs), markTimeBy = TRUE)


# Network measures
igraph::average.path.length(g1)
igraph::transitivity(g1)
qgraph::smallworldness(g1)

recs1 <- igraph::centr_degree(g1)

(RP_N <- sum(recs1$res))
rqa1$RP_N

RP_N / recs1$theoretical_max
rqa1$RR
```

### TS 2 {-}

```{r message=FALSE, warning=FALSE}
#----------------------
# Adjacency matrix TS_2
#----------------------
plot(ts(series$TS_2))

est_parameters(series$TS_2)

RN2 <- rn(y1 = series$TS_2, emDim = 7, emLag = 2, emRad = NA,  targetValue = 0.01)
rn_plot(RN2)

# Get RQA measures
rqa2 <- rp_measures(RN2)

# knitr::kable(rqa2,digits = 2) %>%
#   kable_styling() %>%
#   scroll_box(width = "100%", height = "100px")

g2 <- graph_from_adjacency_matrix(RN2, mode="undirected", diag = FALSE)
V(g2)$size <- degree(g2)
g2r <- casnet::make_spiral_graph(g2,arcs = arcs,epochColours = getColours(arcs), markTimeBy = TRUE)

# Network measures
average.path.length(g2)
transitivity(g2)
smallworldness(g2)

recs2 <- centr_degree(g2)

(RP_N <- sum(recs2$res))
rqa2$RP_N

RP_N / recs2$theoretical_max
rqa2$RR
```

### TS 3 {-}

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
#----------------------
# Adjacency matrix TS_3
#----------------------
plot(ts(series$TS_3))

est_parameters(series$TS_3)

RN3 <- rn(y1 = series$TS_3, emDim = 7, emLag = 8, emRad = NA, targetValue = 0.01)
rn_plot(RN3)

# Get RQA measures
rqa3 <- rp_measures(RN3)

# knitr::kable(rqa3,digits = 2) %>%
#   kable_styling() %>%
#   scroll_box(width = "100%", height = "100px")

g3 <- graph_from_adjacency_matrix(RN3, mode="undirected", diag = FALSE)
V(g3)$size <- degree(g3)
g3r <- casnet::make_spiral_graph(g3,arcs = arcs ,epochColours = getColours(arcs), markTimeBy = TRUE)


average.path.length(g3)
transitivity(g3)
smallworldness(g3)

recs3 <- centr_degree(g3)

(RP_N <- sum(recs3$res))
rqa3$RP_N

RP_N / recs3$theoretical_max
rqa3$RR
```


## Multiplex Recurrence Networks {.tabset .tabset-fade .tabset-pills}


Consider the three time series to be part of a multi-layer recurrence network.
Common properties of the multiplex network are multi-layer mutual information and edge overlap.

```{r}
# Inter-layer Mutual Information
(interlayer_mi12 <- mi_interlayer(g1,g2))
(interlayer_mi13 <- mi_interlayer(g1,g3))
(interlayer_mi23 <- mi_interlayer(g2,g3))

# mean I-L MI  
mean(c(interlayer_mi12,interlayer_mi13,interlayer_mi23))

# Edge Overlap
(edge_overlap12 <- length(E(g1 %s% g2)) / (length(E(g1))+length(E(g2))))
(edge_overlap13 <- length(E(g1 %s% g3)) / (length(E(g1))+length(E(g3))))
(edge_overlap23 <- length(E(g2 %s% g3)) / (length(E(g2))+length(E(g3))))

# mean EO 
mean(c(edge_overlap12,interlayer_mi13,interlayer_mi23))

# Overall Edge Overlap
(eo_all  <- length(E(intersection(g1,g2,g3))))
#(eo_mean <- eo_all / (edge_overlap12+edge_overlap13+edge_overlap23))
```
  
  